### 2020-04-02 06b-bisimulation.tex
Update 06b-bisimulation.tex

\begin{definition}[Satisfaction as membership]\label{sec:semantics}
Let  $\upsilon$ be  a transition label, and $F$ be an assertion.
We write $\upsilon \entails F$ (read as the transition label $v$ satisfies the assertion $F$) 
if $\mathit{flat}(\upsilon)\in \llbracket F\rrbracket$, otherwise we write $\upsilon \not\entails F$ and say that $F$ does not hold at $v$. 
\end{definition}

@@

The notion of bio-simulation builds on the above language of assertions to parameterize the induced equivalence on the property of interest. 
Please recall that we have defined the behaviour of the contest in a non determinist way, thus 
at each step, different possible sets of entities can be provided to the system and different sets of reaction can be enabled/disabled. 
Bio-simulation can thus be used to compare the behaviour of different systems that share some of the reactions or entities or also to compare the behaviour of the same set of reaction rules when different contexts are provided.

@@

\begin{definition}[Bio-similarity  $ \sim_F $]
Given an assertion $F$, a \emph{bio-simulation} $\mathbf{R}_F$ that respects $F$ is a binary relation over {\tt link}-calculus processes such that, if $P \mathrel{\mathbf{R}_F} Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xrightarrow{v} P'$ then there exist $w,Q'$ such that $Q  \xrightarrow{w} Q'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xrightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xrightarrow{v} P'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_F $ denote the largest  bio-simulation and we say that $P$ is \emph{bio-similar} to $Q$, with respect to $F$, if $P \sim_F Q$.
\end{definition}

\begin{remark}
An alternative way to look at a bio-simulation that respects $F$ is to define it as an ordinary bisimulation over the transition system labelled over $\{F,\neg F\}$ obtained by transforming each transition $P \xrightarrow{v} P'$ such that $v\entails F$ into $P \xrightarrow{F} P'$ and  each transition $P \xrightarrow{v} P'$ such that $v\not\entails F$ into $P \xrightarrow{\neg F} P'$.
\end{remark}

@@

\begin{definition}[semantics of $G$]
 We define $\llbracket G\rrbracket \subseteq {\cal P}$ by induction on the structure of $G$:
 %(read as the transition label $v$ satisfies the formula $F$) as: 
\[
\begin{array}{ccl@{\hspace{1cm}}ccl}
 \llbracket {\tt t}\rrbracket &\defeq& {\cal P} &  \llbracket G\wedge H\rrbracket &\defeq&  \llbracket G\rrbracket  \cap \llbracket H\rrbracket  \\
\llbracket {\tt f}\rrbracket& \defeq& \emptyset &
  \llbracket G\vee H\rrbracket &\defeq&  \llbracket G\rrbracket  \cup \llbracket H\rrbracket \\
%\item   $ v \entails F_1 \vee F_2$ iff   $v \entails F_1$  $\vee$  $v \entails F_2$.
\end{array}
\]
\[
\begin{array}{lcl}
 \llbracket  \langle \chi \rangle G \rrbracket &\defeq& \{P \in {\cal P}: \exists v,P'.\ P\xrightarrow{v}P' \mbox{ with } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\} \\
 \llbracket [\chi]G \rrbracket &\defeq& \{P \in {\cal P}: \forall v,P'.\ P\xrightarrow{v}P' \mbox{ implies } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\}
\end{array}
\]
We write $P \entails G$ ($P$ satisfies $G$) if and only if $P \in  \llbracket G \rrbracket$.
\end{definition}

@@

\begin{theorem}
$\sim_F\ =\ \equiv_{\mathcal{L}_F}$
\end{theorem}

>>>>

\begin{definition}[Satisfaction as membership]\label{sec:semantics}
Let  $\upsilon$ be  a transition label, and $F$ be an assertion.
We write $\upsilon \entails F$ (read as the transition label $v$ satisfies the assertion $F$) 
if $\mathit{flat}(\upsilon)\in \llbracket F\rrbracket$, otherwise we write $\upsilon \not\entails F$ (or also $\upsilon \entails \neg F$) and say that $F$ does not hold at $v$. 
\end{definition}

@@

The notion of bio-simulation builds on the above language of assertions to parameterize the induced equivalence on the property of interest. 
Please recall that we have defined the behaviour of the contest in a non deterministic way, thus 
at each step, different possible sets of entities can be provided to the system and different sets of reaction can be enabled/disabled. 
Bio-simulation can thus be used to compare the behaviour of different systems that share some of the reactions or entities or also to compare the behaviour of the same set of reaction rules when different contexts are provided.

@@

\begin{definition}[Bio-similarity  $ \sim_F $]
Given an assertion $F$, a \emph{bio-simulation} $\mathbf{R}_F$ that respects $F$ is a binary relation over c\CNA \  processes such that, whenever $P~\mathrel{\mathbf{R}_F}~Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xRightarrow{v} P'$ then there exist $w,Q'$ such that $Q  \xRightarrow{w} Q'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xRightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xRightarrow{v} P'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_F $ denote the largest  bio-simulation and we say that $P$ is \emph{bio-similar} to $Q$, with respect to $F$, if $P \sim_F Q$.
\end{definition}

\begin{remark}
Please remember that the notation $P \xRightarrow{v} P'$ refers to ordinary transitions $P \xrightarrow{v} P'$ where $v$ is a complete chain (solid and with $\silent$ actions at the extremes).
It should not be confused with the notation for weak transitions commonly found in the literature on process algebras.
\end{remark}

\begin{remark}
An alternative way to look at a bio-simulation that respects $F$ is to define it as an ordinary bisimulation over the transition system labelled over $\{F,\neg F\}$ obtained by transforming each transition $P \xRightarrow{v} P'$ such that $v\entails F$ into $P \xrightarrow{F} P'$ and  each transition $P \xRightarrow{v} P'$ such that $v\not\entails F$ into $P \xrightarrow{\neg F} P'$.
\end{remark}

It can be easily shown that the identity relation is a bio-simulation and that bio-simulations are closed under (relational) inverse, composition and union and that, as a consequence, bio-similarity is an equivalence relation.

@@

\begin{definition}[semantics of $G$]
 We define $\llbracket G\rrbracket \subseteq {\cal P}$ by induction on the structure of $G$:
 %(read as the transition label $v$ satisfies the formula $F$) as: 
\[
\begin{array}{ccl@{\hspace{1cm}}ccl}
 \llbracket {\tt t}\rrbracket &\defeq& {\cal P} &  \llbracket G\wedge H\rrbracket &\defeq&  \llbracket G\rrbracket  \cap \llbracket H\rrbracket  \\
\llbracket {\tt f}\rrbracket& \defeq& \emptyset &
  \llbracket G\vee H\rrbracket &\defeq&  \llbracket G\rrbracket  \cup \llbracket H\rrbracket \\
%\item   $ v \entails F_1 \vee F_2$ iff   $v \entails F_1$  $\vee$  $v \entails F_2$.
\end{array}
\]
\[
\begin{array}{lcl}
 \llbracket  \langle \chi \rangle G \rrbracket &\defeq& \{P \in {\cal P}: \exists v,P'.\ P\xRightarrow{v}P' \mbox{ with } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\} \\
 \llbracket [\chi]G \rrbracket &\defeq& \{P \in {\cal P}: \forall v,P'.\ P\xRightarrow{v}P' \mbox{ implies } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\}
\end{array}
\]
We write $P \entails G$ ($P$ satisfies $G$) if and only if $P \in  \llbracket G \rrbracket$.
\end{definition}

Negation is not included in the syntax, but the converse $\overline{G}$ of a bioHML formula $G$ can be easily defined inductively in the same way as for HML logic.

\begin{definition}[converse]
Given a bioHML formula $G$ we define its converse $\overline{G}$ as follows:
\[
\begin{array}{rcl@{\hspace{1cm}}rcl@{\hspace{1cm}}rcl}
\overline{\tt t} & \defeq & {\tt f} &
\overline{G \wedge H} & \defeq & \overline{G} \vee \overline{H} &
\overline{\langle \chi \rangle G} & \defeq & [\chi]\overline{G}
\\
\overline{\tt f} & \defeq & {\tt t} &
\overline{G \vee H} & \defeq & \overline{G} \wedge \overline{H} &
\overline{[\chi]G} & \defeq & \langle \chi \rangle\overline{G}
\end{array}
\]
\end{definition}

We observe that, as expected, for any bioHML formula $G$ and process $P$ we have $\overline{\overline{G}} = G$ and  $P \entails \overline{G}$ iff  $P\not\entails G$.

@@

\begin{theorem}
$\sim_F\ =\ \equiv_{\mathcal{L}_F}$
\end{theorem}
\begin{proof}
The proof is just an adaptation of the classical result. The two implications are proved separately.
\begin{description}
\item[$\sim_F\ \subseteq\ \equiv_{\mathcal{L}_F}$:]
Given any two processes $P\sim_F Q$ we need to prove that for any bioHML formula $G$ we have $P\entails G$ iff $Q\entails G$. 
Without loss of generality, we prove that $P\entails G$ implies $Q\entails G$.
The proof is by structural induction on $G$.
\begin{itemize}
\item 
if $G={\tt t}$, then $Q\entails G$.

\item
if $G={\tt f}$, then the assumption $P\entails G$ is false and the implication holds.

\item
if $G=G_1 \wedge G_2$ we take as inductive hypotheses that
$$
\begin{array}{l}
\forall R,S.\ R \sim_F S \wedge R\entails G_1 \Rightarrow S \entails G_1\\
\forall R,S.\ R \sim_F S \wedge R\entails G_2 \Rightarrow S \entails G_2
\end{array}
$$
We need to prove that $Q\entails G$.
Since $P\entails G=G_1 \wedge G_2$ we have $P\entails G_1$ and $P\entails G_2$.
Since $P\sim_F Q$, by inductive hypotheses we get $Q\entails G_1$ and $Q\entails G_2$.
Hence $Q\entails G_1 \wedge G_2=G$.

\item
if $G=G_1 \vee G_2$  we take as inductive hypotheses that
$$
\begin{array}{l}
\forall R,S.\ R \sim_F S \wedge R\entails G_1 \Rightarrow S \entails G_1\\
\forall R,S.\ R \sim_F S \wedge R\entails G_2 \Rightarrow S \entails G_2
\end{array}
$$
We need to prove that $Q\entails G$.
Since $P\entails G=G_1 \vee G_2$ we have $P\entails G_1$ or $P\entails G_2$.
If $P\entails G_1$, since $P\sim_F Q$, by inductive hypotheses we get $Q\entails G_1$ and thus $Q\entails G_1 \vee G_2=G$.
If $P\entails G_2$, since $P\sim_F Q$, by inductive hypotheses we get $Q\entails G_2$ and thus $Q\entails G_1 \vee G_2=G$.

\item
if $G=\langle\chi\rangle H$ we take as inductive hypothesis that
$$
\forall R,S.\ R \sim_F S \wedge R\entails H \Rightarrow S \entails H
$$
We need to prove that $Q\entails G$.
Since $P\entails \langle\chi\rangle H$ it means that there exists $\upsilon, P'$ such that $P\xRightarrow{\upsilon} P'$ with $\upsilon\entails \chi$ and $P'\entails H$.
Since $P\sim_F Q$, there exists $w,Q'$ such that $Q\xRightarrow{w} Q'$ with $w\entails \chi$ and $P'\sim_F Q'$. Then, by inductive hypothesis, $Q'\entails H$ and thus $Q\entails \langle\chi\rangle H=G$.

\item
if $G=[\chi] H$ we take as inductive hypothesis that
$$
\forall R,S.\ R \sim_F S \wedge R\entails H \Rightarrow S \entails H
$$
We need to prove that $Q\entails G$.
If there is no $\upsilon\entails \chi$ such that $Q\xRightarrow{\upsilon}Q'$ for some $Q'$ then 
 $Q\entails [\chi] H=G$ trivially.
For any $\upsilon,Q'$ such that $Q\xRightarrow{\upsilon}Q'$ with $\upsilon\entails \chi$, then as $P\sim_F Q$ there must exist $w,P'$ such that $P\xRightarrow{w}P'$ with $w\entails \chi$ and $P'\sim_F Q'$. Since $P\entails G=[\chi] H$ then it must be $P'\entails H$. Since $P'\sim_F Q'$, by inductive hypothesis $Q'\entails H$. Hence $Q\entails [\chi] H = G$.
\end{itemize}

\item[$\equiv_{\mathcal{L}_F}\ \subseteq\ \sim_F$:]
We prove that $\equiv_{\mathcal{L}_F}$ is a bio-simulation and thus included in $\sim_F$.
Take two generic processes $P\equiv_{\mathcal{L}_F} Q$ and suppose $P\xRightarrow{\upsilon}P'$ for some $\upsilon,P'$. 
\begin{itemize}
\item
If $\upsilon \entails F$ we want to prove that there exists some $w,Q'$ such that $Q\xRightarrow{w}Q'$, with $w\entails F$ and $P'\equiv_{\mathcal{L}_F} Q'$.

Towards a contradiction, assume that we cannot find such $w,Q'$.

If there is no transition $Q\xRightarrow{w}Q'$ such that $w\entails F$, then the bioHML formula $G \defeq \langle F\rangle{\tt t}$ is such that $P\entails G$ and $Q\not\entails G$, contradicting the assumption $P\equiv_{\mathcal{L}_F} Q$.

Otherwise, let $\mathcal{Q}\defeq \{Q'~\mid~\exists w.\ Q\xRightarrow{w}Q' \ \wedge\ w\entails F\}$ be the set of processes reachable from $Q$ via a transition with a (complete) label that satisfies $F$. Since our processes are with guarded recursion, the set $\mathcal{Q}$ is finite. Let $\mathcal{Q} = \{Q'_1,...,Q'_n\}$. By hypothesis all processes in $\mathcal{Q}$ must not be bio-logically equivalent to $P'$, hence for any $i\in[1,n]$ there exists a bioHML formula $G_i$ such that $P'\entails G_i$ and $Q'_i\not\entails G_i$ (if it was the opposite, $P'\not\entails G_i$ and $Q'_i\entails G_i$, we can use the converse formulas). But then the formula $G\defeq \langle F\rangle(G_1\wedge\dots\wedge G_n)$ is such that $P\entails G$ and $Q\not\entails G$, contradicting the assumption $P\equiv_{\mathcal{L}_F} Q$.

\item
If $\upsilon \not\entails F$ then the proof is analogous to the previous case and thus omitted.
\end{itemize}
\end{description}
\end{proof}
